---
title: "UGS_pipeline"
format: html
---
# =====================================================
# Socioeconomic Segregation and Park Greenness – Reproducible analysis (PAPER)
# =====================================================
This script reproduces the analyses reported in the manuscript.
It uses final, processed datasets and follows the Results structure.

# -----------------------------------------------------
# 0. Setup
# -----------------------------------------------------

# 0.1 Packages
```{r}
#| message: false
#| warning: false
#| include: false

pckgs <- c(
  "here", "dplyr", "tidyr", "readr", "readxl", "stringr", "forcats", "purrr",
  "ggplot2", "sf", "terra", "tmap", "janitor", "broom"
)

invisible(lapply(pckgs, require, character.only = TRUE))
```

# 0.2 Global options
```{r}
options(
  scipen = 999,
  dplyr.summarise.inform = FALSE
)

sf::sf_use_s2(FALSE)
```

# 0.3 Paths
```{r}
dir_data <- here::here("data")

dir_figures <- here::here("outputs", "figures")
dir_tables <- here::here("outputs", "tables")
dir_supplementary <- here::here("outputs", "supplementary")

dirs <- c(
  dir_figures,
  dir_tables,
  dir_supplementary
)

invisible(lapply(
  dirs,
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
))
```


# 0.4 Constants
```{r}
city_levels <- c(
  "Antofagasta",
  "La Serena",
  "Valparaíso",
  "Santiago",
  "Concepción",
  "Puerto Montt"
)

as_city_factor <- function(x) {
  factor(x, levels = city_levels, ordered = TRUE)
}

city_recode <- c(
  "Coquimbo-La Serena" = "La Serena",
  "Gran Valparaíso" = "Valparaíso",
  "Gran Santiago" = "Santiago",
  "Gran Concepción" = "Concepción"
)

ismt_levels <- c(
  "Low",
  "Medium Low",
  "Medium High",
  "High"
)

as_ismt_factor <- function(x) {
  factor(x, levels = ismt_levels, ordered = TRUE)
}
```

# 0.5 Plot theme
```{r}
theme_ugs <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      legend.title = element_text(face = "bold")
    )
}
```


# -----------------------------------------------------
# 1. Data loading
# -----------------------------------------------------

# 1.1 Official supplementary layers
```{r}
# Path ----

official_gpkg <- here::here("data", "UGS_cartography.gpkg")

# Available layers ----

sf::st_layers(official_gpkg)

# Load layers ----

parks_official_sf <- sf::st_read(
  official_gpkg,
  layer = "parks",
  quiet = TRUE
)

ismt_official_sf <- sf::st_read(
  official_gpkg,
  layer = "ismt",
  quiet = TRUE
)

voronoi_official_sf <- sf::st_read(
  official_gpkg,
  layer = "voronoi",
  quiet = TRUE
)

# Checks ----

official_layers_check <- tibble::tibble(
  layer = c("parks", "ismt", "voronoi"),
  object = c("parks_official_sf", "ismt_official_sf", "voronoi_official_sf"),
  n_features = c(
    nrow(parks_official_sf),
    nrow(ismt_official_sf),
    nrow(voronoi_official_sf)
  ),
  crs_epsg = c(
    sf::st_crs(parks_official_sf)$epsg,
    sf::st_crs(ismt_official_sf)$epsg,
    sf::st_crs(voronoi_official_sf)$epsg
  )
)

official_layers_check
```

# 1.2 Original raw spatial data
```{r}
# Parks ----

parks_raw_sf <- sf::st_read(
  here::here("data", "CALIDAD_pzpq_2019", "CALIDAD_pzpq_2019.shp"),
  quiet = TRUE
) %>%
  sf::st_transform(32719) %>%
  dplyr::mutate(
    CITY = dplyr::case_when(
      PROVINCIA == "ANTOFAGASTA" ~ "Antofagasta",
      COMUNA %in% c("LA SERENA", "COQUIMBO") ~ "La Serena",
      COMUNA %in% c("VALPARAÍSO", "VIÑA DEL MAR", "QUILPUÉ", "VILLA ALEMANA") ~ "Valparaíso",
      PROVINCIA == "SANTIAGO" |
        COMUNA %in% c(
          "COLINA", "LAMPA", "PADRE HURTADO",
          "PEÑAFLOR", "PIRQUE", "PUENTE ALTO", "SAN BERNARDO"
        ) ~ "Santiago",
      COMUNA %in% c(
        "CONCEPCIÓN", "TALCAHUANO", "HUALPÉN", "PENCO",
        "CORONEL", "SAN PEDRO DE LA PAZ"
      ) ~ "Concepción",
      COMUNA %in% c("PUERTO MONTT", "PUERTO VARAS") ~ "Puerto Montt",
      TRUE ~ NA_character_
    ),
    CITY = as_city_factor(CITY)
  ) %>%
  dplyr::filter(!is.na(CITY))


# ISMT ----

ismt_raw_sf <- sf::st_read(
  here::here("data", "ISMT_2022_actualizado", "ISMT_2022_actualizado.shp"),
  quiet = TRUE
) %>%
  sf::st_transform(32719) %>%
  dplyr::mutate(
    CITY = dplyr::case_when(
      provincia == "ANTOFAGASTA" ~ "Antofagasta",
      provincia == "COQUIMBO" ~ "La Serena",
      comuna %in% c("VALPARAISO", "VINA_DEL_MAR", "QUILPUE", "VILLA_ALEMANA") ~ "Valparaíso",
      provincia == "SANTIAGO" |
        comuna %in% c(
          "COLINA", "LAMPA", "PADRE_HURTADO", "PENAFLOR",
          "PIRQUE", "PUENTE_ALTO", "SAN_BERNARDO"
        ) ~ "Santiago",
      comuna %in% c(
        "CONCEPCION", "TALCAHUANO", "HUALPEN", "PENCO",
        "CORONEL", "SAN_PEDRO_DE_LA_PAZ"
      ) ~ "Concepción",
      comuna %in% c("PUERTO_MONTT", "PUERTO_VARAS") ~ "Puerto Montt",
      TRUE ~ NA_character_
    ),
    CITY = as_city_factor(CITY)
  ) %>%
  dplyr::filter(!is.na(CITY))


# Voronoi ----

voronoi_raw_sf <- sf::st_read(
  here::here("data", "voronoi_chile", "voronoi_ismt_chile.gpkg"),
  quiet = TRUE
) %>%
  sf::st_transform(32719)


# Checks ----

raw_spatial_check <- tibble::tibble(
  object = c("parks_raw_sf", "ismt_raw_sf", "voronoi_raw_sf"),
  n_features = c(
    nrow(parks_raw_sf),
    nrow(ismt_raw_sf),
    nrow(voronoi_raw_sf)
  ),
  crs_epsg = c(
    sf::st_crs(parks_raw_sf)$epsg,
    sf::st_crs(ismt_raw_sf)$epsg,
    sf::st_crs(voronoi_raw_sf)$epsg
  )
)

raw_spatial_check
```

# 1.3 Raw mobility data
```{r}
entropy_files <- list.files(
  path = here::here("data", "voronoi_chile"),
  pattern = "^H_.*\\.csv$",
  full.names = TRUE
)

entropy_raw_tbl <- purrr::map_dfr(
  entropy_files,
  ~ readr::read_csv(.x, show_col_types = FALSE) %>%
    janitor::clean_names() %>%
    dplyr::mutate(source_file = basename(.x))
) %>%
  dplyr::mutate(
    CITY = dplyr::case_when(
      source_file == "H_antofa.csv" ~ "Antofagasta",
      source_file == "H_laserena.csv" ~ "La Serena",
      source_file == "H_valpo.csv" ~ "Valparaíso",
      source_file == "H_stgo.csv" ~ "Santiago",
      source_file == "H_conce.csv" ~ "Concepción",
      source_file == "H_ptomontt.csv" ~ "Puerto Montt",
      TRUE ~ NA_character_
    ),
    CITY = as_city_factor(CITY)
  )

# Checks -----

entropy_raw_tbl %>%
  dplyr::count(CITY)

entropy_raw_tbl %>%
  dplyr::summarise(
    rows = dplyr::n(),
    antennas = dplyr::n_distinct(antena),
    hours = dplyr::n_distinct(hour)
  )
```


# 1.4 NDVI rasters
```{r}
ndvi_files <- list(
  "Antofagasta"  = "ndvi_antofa_feb2016.tif",
  "La Serena"   = "ndvi_laserena_feb2016.tif",
  "Valparaíso"  = "ndvi_valparaiso_feb2016.tif",
  "Santiago"    = "ndvi_santiago_mar2016.tif",
  "Concepción"  = "ndvi_conce_feb2016.tif",
  "Puerto Montt" = "ndvi_ptomontt_mar2016.tif"
)

ndvi_rasters <- purrr::map(
  ndvi_files,
  ~ raster::raster(here::here("data", "NDVI_rasters_2016", .x))
)

ndvi_rasters <- purrr::map(
  ndvi_rasters,
  ~ {
    r <- .x
    r[r < 0] <- NA
    r
  }
)

# Checks
purrr::imap_dfr(
  ndvi_rasters,
  ~ tibble::tibble(
    CITY = .y,
    min_ndvi = raster::cellStats(.x, "min"),
    max_ndvi = raster::cellStats(.x, "max"),
    mean_ndvi = raster::cellStats(.x, "mean"),
    n_cells = raster::ncell(.x),
    crs = as.character(raster::crs(.x))
  )
)

```

# 1.5 Official analytical tables
```{r}
parks_official_tbl <- readr::read_csv(
  here::here("data", "Table_parks.csv"),
  show_col_types = FALSE
)

entropy_official_tbl <- readr::read_csv(
  here::here("data", "Table_entropy.csv"),
  show_col_types = FALSE
)

# Checks ----

parks_official_tbl %>%
  dplyr::glimpse()

entropy_official_tbl %>%
  dplyr::glimpse()

parks_official_tbl %>%
  dplyr::count(CITY) %>%
  print(n = Inf)

entropy_official_tbl %>%
  dplyr::count(CITY) %>%
  print(n = Inf)
```


# 1.6 Ancillary data
```{r}
# Population / density data ----

population_raw_tbl <- readxl::read_excel(
  here::here("data", "poblaciones.xlsx")
) %>%
  janitor::clean_names()


# Urban polygons ----

urban_raw_sf <- sf::st_read(
  here::here("data", "Mapa_Ciudades_2017", "Ciudades_2017.shp"),
  quiet = TRUE
)


# WorldClim temperature rasters ----

tavg_files <- list.files(
  here::here("data", "worldclim", "tavg"),
  pattern = "\\.tif$",
  full.names = TRUE
)

tavg_raw_rasters <- terra::rast(tavg_files)


# WorldClim precipitation rasters ----

prec_files <- list.files(
  here::here("data", "worldclim", "prec"),
  pattern = "\\.tif$",
  full.names = TRUE
)

prec_raw_rasters <- terra::rast(prec_files)


# Checks ----

ancillary_check <- tibble::tibble(
  object = c(
    "population_raw_tbl",
    "urban_raw_sf",
    "tavg_raw_rasters",
    "prec_raw_rasters"
  ),
  n = c(
    nrow(population_raw_tbl),
    nrow(urban_raw_sf),
    terra::nlyr(tavg_raw_rasters),
    terra::nlyr(prec_raw_rasters)
  )
)

ancillary_check
```


# -----------------------------------------------------
# 2. Data preparation
# -----------------------------------------------------
## 2.1 Prepare park geometries
```{r}
parks_sf <- parks_raw_sf %>%
  mutate(
    ID = as.character(ID_TEXT),
    CITY = as.character(CITY),
    AP = as.numeric(st_area(.))
  ) %>%
  group_by(ID) %>%
  mutate(
    AT = sum(AP, na.rm = TRUE)
  ) %>%
  ungroup()

ismt_sf <- ismt_raw_sf

# Quick checks
nrow(parks_sf)
sum(is.na(parks_sf$ID))
sum(is.na(parks_sf$CITY))
summary(parks_sf$AP)
summary(parks_sf$AT)

# AP represents the area of each individual park polygon.
# Several polygons may share the same park ID.
#
# AT represents the total area of each park (sum of all polygons
# sharing the same ID) and is later used as the denominator for
# area-weighted spatial overlays.
```

```{r}
parks_sf %>%
  st_drop_geometry() %>%
  arrange(desc(AT)) %>%
  select(ID, CITY, NOMBRE_EP, COMUNA, AT) %>%
  head(10)
```

## 2.2 ISMT reconstruction
```{r}
# Intersect parks with ISMT polygons.
# This splits parks that overlap more than one census zone.
parks_ismt_inter <- st_intersection(
  parks_sf,
  ismt_sf
)

# Recalculate partial area after intersection.
# AT represents the original total park area.
parks_ismt_inter <- parks_ismt_inter %>%
  mutate(
    AP_inter = as.numeric(st_area(.)),
    area_weight = AP_inter / AT,
    ismt_weighted = ismtpn * area_weight
  )

# Calculate area-weighted ISMT for each park
parks_ismt_tbl <- parks_ismt_inter %>%
  st_drop_geometry() %>%
  group_by(ID) %>%
  summarise(
    ISMT = sum(ismt_weighted, na.rm = TRUE),
    .groups = "drop"
  )

# Join reconstructed ISMT
parks_sf <- parks_sf %>%
  select(-any_of("ISMT")) %>%
  left_join(parks_ismt_tbl, by = "ID")

# Quick checks
nrow(parks_sf)
sum(is.na(parks_sf$ISMT))
summary(parks_sf$ISMT)
```

###2.2.1 ISMT categorization
```{r}
# Assign city-specific ISMT categories used in the original analysis.
#
# These thresholds were defined in the original workflow and were based on
# the distribution of socioeconomic status associated with mobile-phone users.
# The underlying dataset used to derive these thresholds is not available
# in the current reproducibility package.
#
# Therefore, the original thresholds are reproduced here unchanged to ensure
# full consistency with the published analyses.
#
# NOTE:
# ISMT_LEVEL is not derived from the reconstructed park ISMT distribution.
# Categories reproduce the original city-specific thresholds used in the study.

parks_sf <- parks_sf %>%
  mutate(
    ISMT_LEVEL = case_when(

      CITY == "Antofagasta" ~ as.character(
        cut(
          ISMT,
          breaks = c(0, 0.166461, 0.527227, 0.630018, 1),
          labels = c("Low", "Medium Low", "Medium High", "High"),
          include.lowest = TRUE
        )
      ),

      CITY == "La Serena" ~ as.character(
        cut(
          ISMT,
          breaks = c(0, 0.544723, 0.592560, 0.663751, 1),
          labels = c("Low", "Medium Low", "Medium High", "High"),
          include.lowest = TRUE
        )
      ),

      CITY == "Valparaíso" ~ as.character(
        cut(
          ISMT,
          breaks = c(0, 0.547882, 0.609700, 0.675941, 1),
          labels = c("Low", "Medium Low", "Medium High", "High"),
          include.lowest = TRUE
        )
      ),

      CITY == "Santiago" ~ as.character(
        cut(
          ISMT,
          breaks = c(0, 0.535257, 0.618865, 0.733090, 1),
          labels = c("Low", "Medium Low", "Medium High", "High"),
          include.lowest = TRUE
        )
      ),

      CITY == "Concepción" ~ as.character(
        cut(
          ISMT,
          breaks = c(0, 0.544723, 0.592560, 0.663751, 1),
          labels = c("Low", "Medium Low", "Medium High", "High"),
          include.lowest = TRUE
        )
      ),

      CITY == "Puerto Montt" ~ as.character(
        cut(
          ISMT,
          breaks = c(0, 0.521255, 0.559032, 0.648573, 1),
          labels = c("Low", "Medium Low", "Medium High", "High"),
          include.lowest = TRUE
        )
      ),

      TRUE ~ NA_character_
    ),

    ISMT_LEVEL = as_ismt_factor(ISMT_LEVEL)
  )

# Quick checks
parks_sf %>%
  st_drop_geometry() %>%
  count(CITY, ISMT_LEVEL)
```

### 2.2.2 Validation
```{r}
# ============================================================
# Validation against official analytical table
# ============================================================

table_parks_old_id <- parks_official_tbl %>%
  mutate(
    ID = as.character(ID)
  ) %>%
  group_by(ID) %>%
  summarise(
    ISMT_old = first(na.omit(ISMT)),
    n_old = n(),
    .groups = "drop"
  )

parks_ismt_id <- parks_sf %>%
  st_drop_geometry() %>%
  group_by(ID) %>%
  summarise(
    ISMT_new = first(na.omit(ISMT)),
    n_new = n(),
    .groups = "drop"
  )

check_ismt <- parks_ismt_id %>%
  inner_join(
    table_parks_old_id,
    by = "ID"
  ) %>%
  mutate(
    diff = abs(ISMT_new - ISMT_old)
  )

# Number of matching parks
nrow(check_ismt)

# Correlation between reconstructed and official ISMT
cor(
  check_ismt$ISMT_new,
  check_ismt$ISMT_old,
  use = "complete.obs"
)

# Absolute differences
summary(check_ismt$diff)

# Mean absolute difference
mean(check_ismt$diff, na.rm = TRUE)

# Parks with largest discrepancies
check_ismt %>%
  arrange(desc(diff)) %>%
  select(
    ID,
    ISMT_old,
    ISMT_new,
    diff,
    n_old,
    n_new
  ) %>%
  head(20)

# Validation summary
#
# The original analyses were conducted using city-specific scripts.
# Here, the workflow was reconstructed into a unified reproducible pipeline.
#
# Comparison against the official analytical table:
# - Matching parks: 10,908
# - Pearson correlation (ISMT): ~0.9994
# - Mean absolute difference: ~0.00012
#
# Small discrepancies are expected due to differences in polygon
# fragmentation and geometry handling between the original and
# reconstructed workflows.
```


## 2.3 NDVI reconstruction
```{r}
# ============================================================
# 2.3 NDVI reconstruction
# ============================================================

# Reproject NDVI rasters to the park CRS (EPSG:32719)
# This reproduces the original workflow used in the city-specific scripts.

ndvi_rasters_utm <- purrr::map(
  ndvi_rasters,
  ~ {
    r <- raster::projectRaster(
      .x,
      crs = sf::st_crs(parks_sf)$wkt
    )

    r[r < 0] <- NA

    r
  }
)

# Remove previous NDVI column if present

parks_sf <- parks_sf %>%
  dplyr::select(-dplyr::any_of("mean_ndvi")) %>%
  dplyr::mutate(mean_ndvi = NA_real_)

# Extract mean NDVI by city

for (city in names(ndvi_rasters_utm)) {

  parks_city <- parks_sf %>%
    dplyr::filter(CITY == city)

  parks_sf$mean_ndvi[parks_sf$CITY == city] <-
    raster::extract(
      ndvi_rasters_utm[[city]],
      as(parks_city, "Spatial"),
      fun = mean,
      na.rm = TRUE
    )
}

# Quick checks

nrow(parks_sf)

sum(is.na(parks_sf$mean_ndvi))

summary(parks_sf$mean_ndvi)
```
### 2.3.1 Validation
```{r}
# ============================================================
# 2.3.1 Validation against official analytical table
# ============================================================

table_parks_ndvi <- parks_official_tbl %>%
  dplyr::mutate(
    ID = as.character(ID)
  ) %>%
  dplyr::group_by(ID) %>%
  dplyr::summarise(
    mean_ndvi_old = first(na.omit(MEAN_NDVI)),
    .groups = "drop"
  )

parks_ndvi <- parks_sf %>%
  sf::st_drop_geometry() %>%
  dplyr::group_by(ID) %>%
  dplyr::summarise(
    mean_ndvi_new = first(na.omit(mean_ndvi)),
    .groups = "drop"
  )

check_ndvi <- parks_ndvi %>%
  dplyr::inner_join(
    table_parks_ndvi,
    by = "ID"
  ) %>%
  dplyr::mutate(
    diff = abs(mean_ndvi_new - mean_ndvi_old)
  )

# Matching parks

nrow(check_ndvi)

# Correlation

cor(
  check_ndvi$mean_ndvi_new,
  check_ndvi$mean_ndvi_old,
  use = "complete.obs"
)

# Absolute differences

summary(check_ndvi$diff)

# Mean absolute difference

mean(check_ndvi$diff, na.rm = TRUE)

# Largest discrepancies

check_ndvi %>%
  dplyr::arrange(desc(diff)) %>%
  dplyr::slice_head(n = 20)

# Validation summary:
# - Matching parks: 10,908
# - Pearson correlation (NDVI): 0.9837
# - Mean absolute difference: 0.0028
# - 75% of parks have zero difference
# - 216 parks show differences > 0.05, mostly in Santiago
#
# The remaining discrepancies likely reflect minor differences between the
# original city-specific scripts and the unified reconstruction pipeline.
# Published results are reproduced using parks_official_tbl.
```

## 2.4 Entropy reconstruction
The entropy reconstruction reproduced the expected analytical structure
(park × hour records across the six cities), but did not fully reproduce the
published entropy values. After harmonizing city names, the validation matched
261,589 park-hour records across 10,903 parks, with an overall Pearson
correlation of 0.851 and a mean absolute difference of 0.027. Antofagasta was
reproduced exactly, while the remaining cities showed moderate deviations.

Because the published results are replicated from the official analytical
table (`Table_entropy.csv`), the reconstructed entropy object is retained only
as methodological documentation. The source of the remaining discrepancies is
left as a documented limitation of the reconstruction and should be revisited
by comparing the raw city-specific H/N input files used in the original
analysis.
```{r}
parks_voronoi_inter <- st_intersection(
  voronoi_raw_sf,
  parks_sf
) %>%
  mutate(
    AP_voronoi = as.numeric(st_area(.)),
    area_weight_voronoi = AP_voronoi / AT
  )

parks_entropy_inter <- parks_voronoi_inter %>%
  st_drop_geometry() %>%
  left_join(
    entropy_raw_tbl,
    by = c("ID_OLD" = "antena")
  )

entropy_reconstructed_tbl <- parks_entropy_inter %>%
  filter(!is.na(h)) %>%
  group_by(
    ID = ID.1,
    CITY = CITY.x,
    hour
  ) %>%
  summarise(
    H = sum(h * area_weight_voronoi, na.rm = TRUE),
    N = sum(n * area_weight_voronoi, na.rm = TRUE),
    .groups = "drop"
  )
```

## 2.5 Analytical categorization
```{r}
parks_analysis_tbl <- parks_official_tbl %>%
  mutate(
    ID = as.character(ID),
    CITY = recode(
      CITY,
      "Coquimbo-La Serena" = "La Serena",
      "Gran Valparaíso" = "Valparaíso",
      "Gran Santiago" = "Santiago",
      "Gran Concepción" = "Concepción"
    )
  ) %>%
  group_by(CITY) %>%
  mutate(
    GREENNESS_LEVEL = cut(
      MEAN_NDVI,
      breaks = quantile(
        MEAN_NDVI,
        probs = seq(0, 1, 1/3),
        na.rm = TRUE
      ),
      labels = c(
        "Low Greenness",
        "Medium Greenness",
        "High Greenness"
      ),
      include.lowest = TRUE
    )
  ) %>%
  ungroup()

entropy_analysis_tbl <- entropy_official_tbl %>%
  mutate(
    ID = as.character(ID),
    CITY = recode(
      CITY,
      "Coquimbo-La Serena" = "La Serena",
      "Gran Valparaíso" = "Valparaíso",
      "Gran Santiago" = "Santiago",
      "Gran Concepción" = "Concepción"
    )
  )
```

# -----------------------------------------------------
# 3. Results
# -----------------------------------------------------
# Fig. 2: NDVI distribution by city
```{r}
# Prepare plotting data
fig2_ndvi_data <- parks_analysis_tbl %>%
  mutate(
    CITY = factor(
      CITY,
      levels = c(
        "Antofagasta",
        "La Serena",
        "Valparaíso",
        "Santiago",
        "Concepción",
        "Puerto Montt"
      )
    )
  ) %>%
  filter(
    !is.na(MEAN_NDVI),
    !is.na(CITY)
  )

# Check
fig2_ndvi_data %>%
  count(CITY)

summary(fig2_ndvi_data$MEAN_NDVI)

# Plot
fig2_ndvi_hist <- ggplot(
  fig2_ndvi_data,
  aes(x = MEAN_NDVI)
) +
  geom_histogram(
    bins = 30,
    color = "black",
    fill = "grey75"
  ) +
  facet_wrap(
    ~ CITY,
    ncol = 2,
    scales = "free_y"
  ) +
  labs(
    x = "Mean NDVI",
    y = "Number of parks"
  ) +
  theme_bw() +
  theme(
    strip.background = element_rect(
      fill = "grey90",
      color = "black"
    ),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

fig2_ndvi_hist

# Export
ggsave(
  filename = here("outputs", "figures", "fig2_ndvi_histogram_by_city.png"),
  plot = fig2_ndvi_hist,
  width = 8,
  height = 6,
  dpi = 300
)

ggsave(
  filename = here("outputs", "figures", "fig2_ndvi_histogram_by_city.pdf"),
  plot = fig2_ndvi_hist,
  width = 8,
  height = 6
)
```


#Fig. 3: NDVI x POPULATION
```{r}
# -----------------------------------------------------
# 3.1 Greenness of UGS
# Fig. 3 – City population size and park greenness
# -----------------------------------------------------

# Mean NDVI by city
mean_ndvi_city <- parks_analysis_tbl %>%
  group_by(CITY) %>%
  summarise(
    NDVI_mean = mean(MEAN_NDVI, na.rm = TRUE),
    NDVI_se = sd(MEAN_NDVI, na.rm = TRUE) / sqrt(sum(!is.na(MEAN_NDVI))),
    .groups = "drop"
  )

# Population data
# Population expressed as thousands of inhabitants.
# These values reproduce those used in the original figure script.
population_city_tbl <- tibble::tribble(
  ~CITY,            ~population_thousands,
  "Antofagasta",              361.9,
  "La Serena",                534.2,
  "Valparaíso",               909.2,
  "Santiago",                6254.3,
  "Concepción",              1187.2,
  "Puerto Montt",             308.1
)

# Join data
fig3_population_ndvi_data <- mean_ndvi_city %>%
  left_join(
    population_city_tbl,
    by = "CITY"
  ) %>%
  mutate(
    CITY = factor(
      CITY,
      levels = c(
        "Antofagasta",
        "La Serena",
        "Valparaíso",
        "Santiago",
        "Concepción",
        "Puerto Montt"
      )
    )
  )

fig3_population_ndvi_data

# Export supplementary table
write_csv(
  fig3_population_ndvi_data,
  here("outputs", "tables", "fig3_population_ndvi_summary.csv")
)

# Plot
city_colors <- c(
  "Antofagasta"  = "#66c2a5",
  "La Serena"    = "#fc8d62",
  "Valparaíso"   = "#8da0cb",
  "Concepción"   = "#a6d854",
  "Santiago"     = "#e78ac3",
  "Puerto Montt" = "#d8b60a"
)

fig3_population_ndvi <- ggplot(
  fig3_population_ndvi_data,
  aes(
    x = population_thousands,
    y = NDVI_mean,
    label = CITY,
    color = CITY
  )
) +
  geom_pointrange(
    aes(
      ymin = NDVI_mean - 1.96 * NDVI_se,
      ymax = NDVI_mean + 1.96 * NDVI_se
    ),
    linewidth = 0.7,
    fatten = 4,
    alpha = 0.9
  ) +
  ggrepel::geom_text_repel(
    size = 4,
    show.legend = FALSE
  ) +
  scale_x_log10(
    breaks = c(300, 500, 1000, 2000, 5000),
    labels = c("300", "500", "1,000", "2,000", "5,000")
  ) +
  scale_color_manual(values = city_colors) +
  theme_minimal(base_size = 14) +
  labs(
    x = expression("Population " %*% 10^3 * " - log scale"),
    y = "NDVI"
  ) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.line  = element_line(linewidth = 0.35, colour = "black"),
    axis.ticks = element_line(linewidth = 0.35, colour = "black")
  )

fig3_population_ndvi

# Export figure
ggsave(
  filename = here("outputs", "figures", "fig3_population_ndvi.pdf"),
  plot = fig3_population_ndvi,
  device = cairo_pdf,
  width = 7,
  height = 4.5,
  dpi = 300
)

ggsave(
  filename = here("outputs", "figures", "fig3_population_ndvi.png"),
  plot = fig3_population_ndvi,
  width = 7,
  height = 4.5,
  dpi = 300
)
```


#Fig. 4: NDVI x ISMT category
```{r}
# -----------------------------------------------------
# 3.2 Sociomaterial context
# Fig. 4 + Table A.2 – Greenness by ISMT category
# -----------------------------------------------------

ismt_order <- c(
  "Low",
  "Medium Low",
  "Medium High",
  "High"
)

city_labels <- c(
  "Antofagasta"  = "A Antofagasta",
  "La Serena"    = "B La Serena",
  "Valparaíso"   = "C Valparaíso",
  "Santiago"     = "D Santiago",
  "Concepción"   = "E Concepción",
  "Puerto Montt" = "F Puerto Montt"
)

fig4_ndvi_ismt_data <- parks_analysis_tbl %>%
  filter(
    !is.na(CITY),
    !is.na(MEAN_NDVI),
    !is.na(ISMT_LEVEL)
  ) %>%
  mutate(
    CITY = factor(
      CITY,
      levels = c(
        "Antofagasta",
        "La Serena",
        "Valparaíso",
        "Santiago",
        "Concepción",
        "Puerto Montt"
      )
    ),
    ISMT_LEVEL = factor(
      ISMT_LEVEL,
      levels = ismt_order
    )
  ) %>%
  filter(
    !is.na(CITY),
    !is.na(ISMT_LEVEL)
  )

# Check
fig4_ndvi_ismt_data %>%
  count(CITY, ISMT_LEVEL)

# ANOVA by city
anova_ndvi_ismt_tbl <- fig4_ndvi_ismt_data %>%
  group_by(CITY) %>%
  group_modify(~ {
    mod <- aov(MEAN_NDVI ~ ISMT_LEVEL, data = .x)
    broom::tidy(mod)
  }) %>%
  ungroup()

write_csv(
  anova_ndvi_ismt_tbl,
  here("outputs", "tables", "fig4_anova_ndvi_ismt_by_city.csv")
)

# Table A.2 – Tukey HSD by city
tukey_ndvi_ismt_tbl <- fig4_ndvi_ismt_data %>%
  group_by(CITY) %>%
  group_modify(~ {
    mod <- aov(MEAN_NDVI ~ ISMT_LEVEL, data = .x)
    tk <- TukeyHSD(mod)$ISMT_LEVEL

    as_tibble(tk, rownames = "comparison")
  }) %>%
  ungroup() %>%
  rename(
    lower = lwr,
    upper = upr,
    p_adj = `p adj`
  ) %>%
  mutate(
    significant = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  )

write_csv(
  tukey_ndvi_ismt_tbl,
  here("outputs", "tables", "table_a2_tukey_ndvi_ismt_by_city.csv")
)

# Plot
fig4_ndvi_ismt <- ggplot(
  fig4_ndvi_ismt_data,
  aes(
    x = ISMT_LEVEL,
    y = MEAN_NDVI,
    colour = ISMT_LEVEL
  )
) +
  geom_boxplot(
    outlier.alpha = 0.35,
    width = 0.65
  ) +
  facet_wrap(
    ~ CITY,
    ncol = 2,
    labeller = labeller(CITY = city_labels)
  ) +
  labs(
    x = "ISMT category",
    y = "Mean NDVI"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 35, hjust = 1),
    panel.grid.minor = element_blank()
  )

fig4_ndvi_ismt

# Export figure
ggsave(
  filename = here("outputs", "figures", "fig4_ndvi_ismt_boxplots.pdf"),
  plot = fig4_ndvi_ismt,
  device = cairo_pdf,
  width = 6,
  height = 10,
  dpi = 300
)

ggsave(
  filename = here("outputs", "figures", "fig4_ndvi_ismt_boxplots.png"),
  plot = fig4_ndvi_ismt,
  width = 6,
  height = 10,
  dpi = 300
)
```




#Fig. 5: Sankey diagrams
```{r}
# Sankey data for all cities
sankey_greenness_ismt_links <- parks_analysis_tbl %>%
  filter(
    !is.na(CITY),
    !is.na(GREENNESS_LEVEL),
    !is.na(ISMT_LEVEL)
  ) %>%
  count(
    CITY,
    source = GREENNESS_LEVEL,
    target = ISMT_LEVEL,
    name = "value"
  )

write_csv(
  sankey_greenness_ismt_links,
  here("outputs", "tables", "sankey_greenness_ismt_links.csv")
)

# Plot one example city
fig5_sankey_antofagasta <- make_sankey_greenness_ismt(
  data = parks_analysis_tbl,
  city_name = "Antofagasta"
)

fig5_sankey_antofagasta

# NOTE:
# The full Sankey flow data are exported for all cities.
# One city is displayed as an example because networkD3 outputs are
# interactive HTML widgets and their layout is generated automatically.
```

#Fig. 6: Timelines
```{r}
entropy_analysis_tbl <- entropy_official_tbl %>%
  mutate(
    CITY = recode(
      CITY,
      "Coquimbo-La Serena" = "La Serena",
      "Gran Valparaíso" = "Valparaíso",
      "Gran Santiago" = "Santiago",
      "Gran Concepción" = "Concepción"
    )
  )
```

```{r}
# -----------------------------------------------------
# 3.3 Segregation of users
# Fig. 6 – Entropy timelines
# -----------------------------------------------------

entropy_long_tbl <- entropy_analysis_tbl %>%
  pivot_longer(
    cols = `0`:`23`,
    names_to = "hour",
    values_to = "H"
  ) %>%
  mutate(hour = as.integer(hour))

entropy_parks_tbl <- parks_analysis_tbl %>%
  select(ID, CITY, GREENNESS_LEVEL, ISMT_LEVEL) %>%
  distinct()

entropy_timeline_data <- entropy_long_tbl %>%
  left_join(
    entropy_parks_tbl,
    by = c("ID", "CITY")
  ) %>%
  filter(
    !is.na(H),
    !is.na(GREENNESS_LEVEL),
    !is.na(ISMT_LEVEL)
  )

timeline_greenness_tbl <- entropy_timeline_data %>%
  group_by(CITY, hour, GREENNESS_LEVEL) %>%
  summarise(
    mean_H = mean(H, na.rm = TRUE),
    .groups = "drop"
  )

timeline_ismt_tbl <- entropy_timeline_data %>%
  group_by(CITY, hour, ISMT_LEVEL) %>%
  summarise(
    mean_H = mean(H, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# -----------------------------------------------------
# 3.3 Segregation of users
# Fig. 6 – Entropy timelines
# -----------------------------------------------------

entropy_long_tbl <- entropy_analysis_tbl %>%
  pivot_longer(
    cols = `0`:`23`,
    names_to = "hour",
    values_to = "H"
  ) %>%
  mutate(hour = as.integer(hour))

entropy_parks_tbl <- parks_analysis_tbl %>%
  select(ID, CITY, GREENNESS_LEVEL, ISMT_LEVEL) %>%
  distinct()

entropy_timeline_data <- entropy_long_tbl %>%
  left_join(
    entropy_parks_tbl,
    by = c("ID", "CITY")
  ) %>%
  filter(
    !is.na(H),
    !is.na(GREENNESS_LEVEL),
    !is.na(ISMT_LEVEL)
  )

timeline_greenness_tbl <- entropy_timeline_data %>%
  group_by(CITY, hour, GREENNESS_LEVEL) %>%
  summarise(
    mean_H = mean(H, na.rm = TRUE),
    .groups = "drop"
  )

timeline_ismt_tbl <- entropy_timeline_data %>%
  group_by(CITY, hour, ISMT_LEVEL) %>%
  summarise(
    mean_H = mean(H, na.rm = TRUE),
    .groups = "drop"
  )
```
```{r}
# Main figure – entropy timeline by greenness, all cities
fig6_timeline_greenness_all <- ggplot(
  timeline_greenness_tbl,
  aes(x = hour, y = mean_H, color = GREENNESS_LEVEL)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.5) +
  facet_wrap(~ CITY, ncol = 2) +
  scale_color_manual(
    values = c(
      "Low Greenness" = "#AFBDAF",
      "Medium Greenness" = "#72AE72",
      "High Greenness" = "#009900"
    )
  ) +
  labs(
    x = "Hour",
    y = "Mean entropy (H)",
    color = "Park greenness"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

fig6_timeline_greenness_all

# ggsave(
#   filename = here("outputs", "figures", "fig6_timeline_greenness_all_cities.pdf"),
#   plot = fig6_timeline_greenness_all,
#   width = 10,
#   height = 7,
#   dpi = 300
# )
# 
# ggsave(
#   filename = here("outputs", "figures", "fig6_timeline_greenness_all_cities.png"),
#   plot = fig6_timeline_greenness_all,
#   width = 10,
#   height = 7,
#   dpi = 300
# )

# NOTE:
# Timelines are reconstructed from the official analytical tables.
# In Antofagasta and Puerto Montt, the vertical ordering of greenness
# categories differs from earlier exploratory plots, likely due to
# differences between the final official tables and the original
# city-specific working objects.
```


```{r}
fig6_timeline_ismt_all <- ggplot(
  timeline_ismt_tbl,
  aes(x = hour, y = mean_H, color = ISMT_LEVEL)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.5) +
  facet_wrap(~ CITY, ncol = 2) +
  labs(
    x = "Hour",
    y = "Mean entropy (H)",
    color = "ISMT category"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

# fig6_timeline_ismt_all
```



#Fig. 7: Cluster analysis
```{r}
# -----------------------------------------------------
# Cluster analysis of park segregation profiles
# -----------------------------------------------------

park_area_tbl <- parks_sf %>%
  st_drop_geometry() %>%
  mutate(
    ID = as.character(ID),
    CITY = as.character(CITY)
  ) %>%
  group_by(ID, CITY) %>%
  summarise(
    AT = max(AT, na.rm = TRUE),
    .groups = "drop"
  )

parks_cluster_tbl <- parks_analysis_tbl %>%
  mutate(
    ID = as.character(ID),
    CITY = as.character(CITY)
  ) %>%
  left_join(
    park_area_tbl %>%
      select(ID, AT),
    by = "ID"
  )

entropy_cluster_tbl <- entropy_analysis_tbl %>%
  mutate(
    ID = as.character(ID),
    CITY = as.character(CITY)
  ) %>%
  left_join(
    parks_cluster_tbl %>%
      select(ID, CITY, AT, QUALITY_INDEX, MEAN_NDVI, ISMT),
    by = c("ID", "CITY")
  ) %>%
  filter(
    AT > 5000
  )
```

```{r}
city_i <- "Antofagasta"

dend_tbl <- entropy_cluster_tbl %>%
  filter(CITY == city_i) %>%
  select(ID, `0`:`23`) %>%
  distinct() %>%
  drop_na()

dend <- dend_tbl %>%
  column_to_rownames("ID") %>%
  as.matrix()

dendro <- as.dendrogram(
  hclust(dist(dend))
)

dendro_colores <- color_branches(dendro, 4) %>%
  set("branches_lwd", 2)

plot(dendro_colores, horiz = TRUE, leaflab = "none")

# The same workflow can be repeated for any city by changing
# the value of `city_i`.
```

```{r}
d4 <- data.frame(cutree(dendro, 4)) %>%
  tibble::rownames_to_column("ID") %>%
  rename(dend_group = cutree.dendro..4.) %>%
  mutate(
    ID = as.character(ID),
    dend_group = factor(dend_group)
  ) %>%
  left_join(
    parks_cluster_tbl %>%
      filter(CITY == city_i) %>%
      select(ID, CITY, AT, QUALITY_INDEX, MEAN_NDVI, ISMT),
    by = "ID"
  ) %>%
  filter(QUALITY_INDEX > 0)
```

```{r}
ggplot(d4, aes(x = dend_group, y = QUALITY_INDEX)) +
  geom_boxplot() +
  theme_minimal()

ggplot(d4, aes(x = dend_group, y = MEAN_NDVI)) +
  geom_boxplot() +
  theme_minimal()

ggplot(d4, aes(x = dend_group, y = ISMT)) +
  geom_boxplot() +
  theme_minimal()
```


# -----------------------------------------------------
# 4. Table 1 – Study area descriptors
# -----------------------------------------------------
```{r}
# =====================================================
# Table 1 – Final study area descriptors
# =====================================================
# The final Table 1 in the manuscript combines descriptors
# from several sources:
#
# - Number of parks, total park area, mean park area, and NDVI:
#   derived from parks_analysis_tbl.
# - Antennas and SES user counts:
#   derived from the mobility dataset / entropy inputs.
# - Population and population density:
#   derived from census-based population tables.
# - Mean temperature and precipitation:
#   extracted from WorldClim rasters using urban polygons.
# - ISMT:
#   derived from the reconstructed/official ISMT attributes.
#
# Because the published table combines analytical outputs,
# census summaries, climate rasters, and manually formatted ranges,
# the exact manuscript table was assembled manually.
# The code above documents the reproducible derivation of its
# main quantitative components.
```


# 4.1 Population density
```{r}
# =====================================================
# Table 1 – Population density
# =====================================================

library(tidyverse)
library(readxl)
library(here)
library(janitor)

# -----------------------------------------------------
# Load data
# -----------------------------------------------------

density_raw <- read_excel(
  here("data", "poblaciones.xlsx")
) %>%
  clean_names()

# -----------------------------------------------------
# Prepare data
# -----------------------------------------------------

density_tbl <- density_raw %>%
  mutate(
    area_km2 = poblacion / densidad_reportada,

    city = case_when(
      comuna == "antofagasta" ~ "Antofagasta",

      comuna %in% c(
        "la serena",
        "coquimbo"
      ) ~ "La Serena",

      comuna %in% c(
        "valparaiso",
        "viña del mar",
        "quilpue",
        "villa alemana"
      ) ~ "Valparaíso",

      comuna %in% c(
        "santiago (provincia)",
        "colina",
        "lampa",
        "padre hurtado",
        "peñaflor",
        "pirque",
        "puente alto",
        "san bernardo"
      ) ~ "Santiago",

      comuna %in% c(
        "concepcion",
        "talcahuano",
        "hualpen",
        "penco",
        "coronel",
        "san pedro de la paz"
      ) ~ "Concepción",

      comuna %in% c(
        "puerto montt",
        "puerto varas"
      ) ~ "Puerto Montt",

      TRUE ~ NA_character_
    )
  )

# -----------------------------------------------------
# Aggregate by city
# -----------------------------------------------------

city_density <- density_tbl %>%
  group_by(city) %>%
  summarise(
    population = sum(poblacion, na.rm = TRUE),
    area_km2 = sum(area_km2, na.rm = TRUE),
    population_density = population / area_km2,
    .groups = "drop"
  ) %>%
  arrange(
    factor(
      city,
      levels = c(
        "Antofagasta",
        "La Serena",
        "Valparaíso",
        "Santiago",
        "Concepción",
        "Puerto Montt"
      )
    )
  )

city_density
# Population density was calculated using total population and total municipal area of the communes composing each conurbation.
```


```{r}
ciudades <- st_read(
  here("data", "Mapa_Ciudades_2017", "Ciudades_2017.shp")
)

names(ciudades)

nrow(ciudades)
unique(ciudades$URBANO)
```

```{r}
# =====================================================
# Prepare urban polygons for climate extraction
# =====================================================

library(sf)
library(dplyr)
library(here)

ciudades <- st_read(
  here("data", "Mapa_Ciudades_2017", "Ciudades_2017.shp")
)

# -----------------------------------------------------
# 1. Assign urban polygons to study cities
# -----------------------------------------------------

ciudades_study <- ciudades %>%
  mutate(
    CITY = case_when(
      NOM_COMUNA == "ANTOFAGASTA" ~ "Antofagasta",

      NOM_COMUNA %in% c(
        "LA SERENA",
        "COQUIMBO"
      ) ~ "La Serena",

      NOM_COMUNA %in% c(
        "VALPARAÍSO",
        "VIÑA DEL MAR",
        "QUILPUÉ",
        "VILLA ALEMANA"
      ) ~ "Valparaíso",

      NOM_COMUNA %in% c(
        "CERRILLOS", "CERRO NAVIA", "CONCHALÍ", "EL BOSQUE",
        "ESTACIÓN CENTRAL", "HUECHURABA", "INDEPENDENCIA",
        "LA CISTERNA", "LA FLORIDA", "LA GRANJA", "LA PINTANA",
        "LA REINA", "LAS CONDES", "LO BARNECHEA", "LO ESPEJO",
        "LO PRADO", "MACUL", "MAIPÚ", "ÑUÑOA",
        "PEDRO AGUIRRE CERDA", "PEÑALOLÉN", "PROVIDENCIA",
        "PUDAHUEL", "QULICURA", "QUINTA NORMAL", "RECOLETA",
        "RENCA", "SAN JOAQUÍN", "SAN MIGUEL", "SAN RAMÓN",
        "SANTIAGO", "VITACURA",
        "COLINA", "LAMPA", "PADRE HURTADO", "PEÑAFLOR",
        "PIRQUE", "PUENTE ALTO", "SAN BERNARDO"
      ) ~ "Santiago",

      NOM_COMUNA %in% c(
        "CONCEPCIÓN",
        "TALCAHUANO",
        "HUALPÉN",
        "PENCO",
        "CORONEL",
        "SAN PEDRO DE LA PAZ"
      ) ~ "Concepción",

      NOM_COMUNA %in% c(
        "PUERTO MONTT",
        "PUERTO VARAS"
      ) ~ "Puerto Montt",

      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(CITY))

# -----------------------------------------------------
# 2. Check selected polygons
# -----------------------------------------------------

ciudades_study %>%
  st_drop_geometry() %>%
  count(CITY, sort = FALSE) %>%
  print(n = Inf)

ciudades_study %>%
  st_drop_geometry() %>%
  select(CITY, NOM_COMUNA, URBANO, TIPO, CATEGORIA, TOT_PERSON) %>%
  arrange(CITY, NOM_COMUNA) %>%
  print(n = Inf)

# -----------------------------------------------------
# 3. Dissolve polygons into six study cities
# -----------------------------------------------------

ciudades_ugs <- ciudades_study %>%
  group_by(CITY) %>%
  summarise(
    population_urban_2017 = sum(TOT_PERSON, na.rm = TRUE),
    urban_area_m2 = sum(as.numeric(st_area(geometry)), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    urban_area_km2 = urban_area_m2 / 1e6,
    urban_density_km2 = population_urban_2017 / urban_area_km2
  )

# -----------------------------------------------------
# 4. Final check
# -----------------------------------------------------

ciudades_ugs %>%
  st_drop_geometry() %>%
  arrange(factor(
    CITY,
    levels = c(
      "Antofagasta",
      "La Serena",
      "Valparaíso",
      "Santiago",
      "Concepción",
      "Puerto Montt"
    )
  ))
```

# 4.2 Climate variables
```{r}
# Temperature: annual mean, coldest month, warmest month
temp_mean <- terra::extract(tavg_annual, vect(ciudades_ugs_wgs84), fun = mean, na.rm = TRUE)

temp_monthly_city <- lapply(1:nlyr(tavg), function(i) {
  terra::extract(tavg[[i]], vect(ciudades_ugs_wgs84), fun = mean, na.rm = TRUE)[,2]
}) %>%
  bind_cols()

temp_min_month <- apply(temp_monthly_city, 1, min, na.rm = TRUE)
temp_max_month <- apply(temp_monthly_city, 1, max, na.rm = TRUE)
```
```{r}
# Precipitation: annual total, driest month, wettest month
prec_mean <- terra::extract(prec_annual, vect(ciudades_ugs_wgs84), fun = mean, na.rm = TRUE)

prec_monthly_city <- lapply(1:nlyr(prec), function(i) {
  terra::extract(prec[[i]], vect(ciudades_ugs_wgs84), fun = mean, na.rm = TRUE)[,2]
}) %>%
  bind_cols()

prec_min_month <- apply(prec_monthly_city, 1, min, na.rm = TRUE)
prec_max_month <- apply(prec_monthly_city, 1, max, na.rm = TRUE)
```


```{r}
climate_city <- ciudades_ugs %>%
  st_drop_geometry() %>%
  select(CITY) %>%
  mutate(
    mean_annual_temperature = temp_mean[,2],
    min_monthly_temperature = temp_min_month,
    max_monthly_temperature = temp_max_month,
    annual_precipitation = prec_mean[,2],
    min_monthly_precipitation = prec_min_month,
    max_monthly_precipitation = prec_max_month
  )

climate_city
```


#4.3 Antennas
```{r}
voronoi_sf %>%
  st_drop_geometry() %>%
  distinct(ID, City) %>%
  count(City)
```

#4.4 Parks NDVI
```{r}
table_1_ndvi_tbl <- parks_analysis_tbl %>%
  mutate(
    ID = as.character(ID),
    CITY = as.character(CITY)
  ) %>%
  group_by(CITY) %>%
  summarise(
    mean_ndvi = mean(MEAN_NDVI, na.rm = TRUE),
    min_ndvi = min(MEAN_NDVI, na.rm = TRUE),
    max_ndvi = max(MEAN_NDVI, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(
    factor(
      CITY,
      levels = c(
        "Antofagasta",
        "La Serena",
        "Valparaíso",
        "Santiago",
        "Concepción",
        "Puerto Montt"
      )
    )
  )

table_1_ndvi_tbl
```

#4.5 ISMT
```{r}
ismt_city <- ismt_official_sf %>%
  st_drop_geometry() %>%
  mutate(
    population = Q1 + Q2 + Q3 + Q4 + Q5
  ) %>%
  group_by(city) %>%
  summarise(
    mean_ismt =
      weighted.mean(
        ismtpn,
        w = population,
        na.rm = TRUE
      ),

    min_ismt = min(ismtpn, na.rm = TRUE),
    max_ismt = max(ismtpn, na.rm = TRUE),

    total_population = sum(population),
    n_zones = n(),

    .groups = "drop"
  )

ismt_city
```



#4.6 Final table
```{r}
# =====================================================
# Table 1 – Park area descriptors from park geometries
# =====================================================

table_1_parks_tbl <- parks_sf %>%
  st_drop_geometry() %>%
  mutate(
    ID = as.character(ID),
    CITY = as.character(CITY)
  ) %>%
  group_by(CITY, ID) %>%
  summarise(
    park_area_m2 = max(AT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(CITY) %>%
  summarise(
    parks = n_distinct(ID),
    total_park_area_ha = sum(park_area_m2, na.rm = TRUE) / 10000,
    mean_park_area_ha = mean(park_area_m2, na.rm = TRUE) / 10000,
    .groups = "drop"
  ) %>%
  arrange(
    factor(
      CITY,
      levels = c(
        "Antofagasta",
        "La Serena",
        "Valparaíso",
        "Santiago",
        "Concepción",
        "Puerto Montt"
      )
    )
  )

table_1_parks_tbl
```



# -----------------------------------------------------
# 5. Appendix
# -----------------------------------------------------
## 5.1 Spatial characterization

The spatial characterization maps included in the Appendix were
prepared in QGIS using the official park, NDVI, ISMT and entropy
layers. Since these maps are purely illustrative and do not involve
additional analyses, their cartographic composition is not reproduced
in this workflow.

##5.2 ANOVA
```{r}
# -----------------------------------------------------
# 5.2 Appendix – ANOVA tables
# -----------------------------------------------------

## ANOVA: NDVI differences across ISMT quartiles

anova_ndvi_tbl <- parks_analysis_tbl %>%
  st_drop_geometry() %>% 
  filter(
    !is.na(CITY),
    !is.na(MEAN_NDVI),
    !is.na(ISMT_LEVEL)
  ) %>%
  mutate(
    CITY = as.character(CITY),
    ISMT_LEVEL = factor(ISMT_LEVEL)
  ) %>%
  group_by(CITY) %>%
  group_modify(~ {
    
    m1 <- lm(MEAN_NDVI ~ ISMT_LEVEL, data = .x)
    aov_tbl <- anova(m1)
    
    tibble(
      term = rownames(aov_tbl),
      df = aov_tbl$Df,
      sum_sq = aov_tbl$`Sum Sq`,
      mean_sq = aov_tbl$`Mean Sq`,
      f_value = aov_tbl$`F value`,
      p_value = aov_tbl$`Pr(>F)`
    )
    
  }) %>%
  ungroup()
```

```{r}
anova_ndvi_tbl_formatted <- anova_ndvi_tbl %>%
  mutate(
    sum_sq = round(sum_sq, 4),
    mean_sq = round(mean_sq, 4),
    f_value = round(f_value, 3),
    p_value = case_when(
      is.na(p_value) ~ NA_character_,
      p_value < 0.001 ~ "<0.001",
      TRUE ~ as.character(round(p_value, 4))
    )
  ) %>%
  rename(
    City = CITY,
    Term = term,
    `Df` = df,
    `Sum Sq` = sum_sq,
    `Mean Sq` = mean_sq,
    `F-value` = f_value,
    `p-value` = p_value
  )

knitr::kable(
  anova_ndvi_tbl_formatted,
  caption = "Appendix Table A1. One-way ANOVA testing differences in park NDVI among ISMT quartiles within each city.",
  align = "llrrrrr"
)
```



##5.3 Tukey
```{r}
# -----------------------------------------------------
# 5.3 Appendix – Tukey post-hoc table in LaTeX format
# -----------------------------------------------------

library(emmeans)
library(stringr)
library(purrr)

ismt_order <- c("Low", "Medium Low", "Medium High", "High")

city_labels <- c(
  "Antofagasta"  = "\\textbf{A} - Antofagasta",
  "La Serena"   = "\\textbf{B} - La Serena",
  "Valparaíso"  = "\\textbf{C} - Valparaíso",
  "Santiago"    = "\\textbf{D} - Santiago",
  "Concepción"  = "\\textbf{E} - Concepción",
  "Puerto Montt" = "\\textbf{F} - Puerto Montt"
)

city_order <- c(
  "Antofagasta",
  "La Serena",
  "Valparaíso",
  "Santiago",
  "Concepción",
  "Puerto Montt"
)

ismt_order <- c(
  "Low",
  "Medium Low",
  "Medium High",
  "High"
)

tukey_latex_tbl <- parks_analysis_tbl %>%
  st_drop_geometry() %>%
  filter(
    !is.na(CITY),
    !is.na(MEAN_NDVI),
    !is.na(ISMT_LEVEL)
  ) %>%
  mutate(
  CITY = factor(as.character(CITY), levels = city_order),
  ISMT_LEVEL = factor(ISMT_LEVEL, levels = ismt_order)
) %>%
filter(!is.na(CITY), !is.na(ISMT_LEVEL)) %>%
  group_by(CITY) %>%
  group_modify(~ {

    if (n_distinct(.x$ISMT_LEVEL) < 2) return(tibble())

    m1 <- lm(MEAN_NDVI ~ ISMT_LEVEL, data = .x)

    emmeans(m1, pairwise ~ ISMT_LEVEL, adjust = "tukey")$contrasts %>%
      summary(infer = c(TRUE, TRUE)) %>%
      as.data.frame() %>%
      tidyr::separate(
        contrast,
        into = c("I", "J"),
        sep = " - ",
        remove = FALSE
      ) %>%
      transmute(
        I,
        J,
        diff = estimate,
        lower = lower.CL,
        upper = upper.CL,
        p_adj = p.value
      )
  }) %>%
  ungroup()
```


```{r}
fmt_num <- function(x) sprintf("%.3f", x)

fmt_sig <- function(diff, p) {
  value <- fmt_num(diff)
  ifelse(
    p < 0.05,
    paste0("\\textbf{", value, "$^*$}"),
    value
  )
}

fmt_group <- function(x, p) {
  ifelse(p < 0.05, paste0("\\textbf{", x, "}"), x)
}

fmt_ci <- function(lower, upper, p) {
  ci <- paste0("(", fmt_num(lower), ", ", fmt_num(upper), ")")
  ifelse(p < 0.05, paste0("\\textbf{", ci, "}"), ci)
}

tukey_latex_lines <- tukey_latex_tbl %>%
  mutate(
    city_label = city_labels[CITY],
    I_fmt = fmt_group(I, p_adj),
    J_fmt = fmt_group(J, p_adj),
    diff_fmt = fmt_sig(diff, p_adj),
    ci_fmt = fmt_ci(lower, upper, p_adj)
  ) %>%
  group_by(CITY) %>%
  group_split() %>%
  map_chr(function(df) {

    rows <- character(nrow(df))

    for (i in seq_len(nrow(df))) {

      city_cell <- ifelse(i == 1, df$city_label[i], "")
      i_cell <- ifelse(i == 1 || df$I[i] != df$I[i - 1], df$I_fmt[i], "")

      rows[i] <- paste0(
        city_cell, " & ",
        i_cell, " & ",
        df$J_fmt[i], " & ",
        df$diff_fmt[i], " & ",
        df$ci_fmt[i], " \\\\"
      )
    }

    paste0(
      paste(rows, collapse = "\n"),
      "\n\\hline"
    )
  })

cat(
paste0(
"\\begin{table}[!htb]
  \\caption{Tukey post-hoc analysis shows the mean NDVI differences between ISMT quartiles. Significant comparisons between groups I and J are highlighted and marked by an asterisk (\\ie $p < 0.05$).\\label{apxTable:TukeyTable}}
  \\footnotesize
  \\centering
  \\begin{tabularx}{\\linewidth}{l l l r r}
  \\\\[0.2mm]
  \\hline
  \\multirow{2}{*}{\\textbf{City}} & & & \\textbf{Mean difference} & 95\\% Confidence Interval \\\\
  & (I) ISMT Quartile & (J) ISMT Quartile & (I - J) $\\Delta$ NDVI & (lower, upper)\\\\
  \\hline
",
paste(tukey_latex_lines, collapse = "\n\n"),
"
  \\hline
  \\end{tabularx}
\\end{table}"
)
)
```



##5.4 Hierarchical clustering
```{r}
city_i <- "Antofagasta"

dend_tbl <- entropy_cluster_tbl %>%
  filter(CITY == city_i) %>%
  select(ID, `0`:`23`) %>%
  distinct() %>%
  drop_na()

dend <- dend_tbl %>%
  column_to_rownames("ID") %>%
  as.matrix()

dendro <- as.dendrogram(
  hclust(dist(dend))
)

dendro_colores <- color_branches(dendro, 4) %>%
  set("branches_lwd", 2)

plot(dendro_colores, horiz = TRUE, leaflab = "none")

# The same workflow can be repeated for any city by changing
# the value of `city_i`.
```




# -----------------------------------------------------
# 6. Supplementary material
# -----------------------------------------------------

## 6.1 Supplementary cartography

The official supplementary cartography is loaded in Section 1.1 from:
data/UGS_cartography.gpkg

It contains three layers:
 - parks
 - ismt
 - voronoi

These layers are used in this workflow for spatial documentation,
validation and reproducibility checks.

The original supplementary cartographic package is not regenerated here,
because the GeoPackage is treated as an official input rather than an
output of the present repository.




## 6.2 Optional local exports
```{r}
#| eval: false
#| include: false

# Optional: create local copies of selected supplementary layers.
# These files are written to outputs/, which is ignored by Git.

if (!dir.exists(here("outputs"))) dir.create(here("outputs"))

optional_gpkg_path <- here("outputs", "UGS_supplementary_layers_local.gpkg")

if (file.exists(optional_gpkg_path)) file.remove(optional_gpkg_path)

write_sf(parks_official_sf, optional_gpkg_path, layer = "parks")
write_sf(ismt_official_sf, optional_gpkg_path, layer = "ismt")
write_sf(voronoi_official_sf, optional_gpkg_path, layer = "voronoi")

st_layers(optional_gpkg_path)
```

